---
title: "1_Preprocessing"
author: "James Choi"
date: "Last compiled on: `r Sys.Date()`"
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.align='center', 
                      tidy.opts=list(width.cutoff=60), tidy=TRUE)
```

```{r libraries}
# Load libraries
require('Seurat')
require('dplyr')
require('ggplot2')
require('SingleCellExperiment')
```

```{r set_directory}
# Directories for data and results
# Setting working directory to 'scripts'
# setwd('..')
results_out <- '../results/1_preprocessing/'
dir.create(results_out)
```

```{r params_utils}
time_cols <- c('SCtrl' = 'grey90',
               'S1d' = 'red',
               'S7d' = 'blue')
```


```{r load_data}
# Load counts matrix
svz <- read.table(
  file = '../data/raw_counts_all_cells.csv',
  sep = ',',
  stringsAsFactors = FALSE,
  row.names = 1,
  header = TRUE
)
# Convert to sparse, round to integers (has decimals for some reason)
svz <- Matrix::Matrix(data = as.matrix(svz), sparse = TRUE)
svz <- round(svz)
dat_sce <- SingleCellExperiment(assay = list('counts' = svz))
svz <- CreateSeuratObject(counts = svz, project = 'SVZ', assay = 'RNA')
```

First plot some sequencing metrics such as UMIs per cell and number of unique
genes detected per cell.


```{r detection_scatter, fig.height=4, fig.width=4.5, fig.cap='Scatter plot of sequencing metrics by cell.'}
qc <- svz@meta.data
qc <- qc[sample(x = nrow(qc), size = nrow(qc)),]
plot(
  x = qc$nCount_RNA,
  y = qc$nFeature_RNA,
  xlab = 'total UMI',
  ylab = '# genes detected',
  main = 'Detection rates by sample',
  sub = 'Black = SCtrl; Red = S1d; Green = S7d',
  col = sapply(
    X = as.character(qc$orig.ident),
    FUN = function(x) switch(x, 'SCtrl'='black','S1d'='red3','S7d'='green')
  )
)
```

From the above, we see that there are a few cells with outlier numbers of
total UMI and genes detected. These are likely to be doublets. Beyond that,
there are no obvious biases based on sample of origin.


Next we examine cell-level QC metrics.

* nCount_RNA = number of UMIs per cell.
* nFeature_RNA = number of unique genes detected per cell.
* percent.mt = percent of UMIs mapping to a mitochondrial protein-encoding gene.
* log10_total_counts = log10(nCount_RNA)


```{r qc_vln, fig.height=3.5, fig.width=6, fig.cap='Violin plot of various per-cell QC metrics.'}
svz <- PercentageFeatureSet(svz, pattern = '^mt-', col.name = 'percent.mt')
svz$log10_total_counts <- log10(svz$nCount_RNA)
qc_plot <- svz@meta.data %>%
  select(-c('log10_total_counts')) %>%
  reshape2::melt(id.vars = 'orig.ident') %>%
  ggplot(mapping = aes(x = orig.ident, y = value)) +
  geom_violin(scale = 'width') +
  ggbeeswarm::geom_quasirandom(size = 1) +
  facet_wrap(. ~ variable, scales = 'free_y') +
  labs(title = 'Quality control metrics') +
  theme(axis.title = element_blank(),
        plot.title = element_text(size = 18, face = 'bold'),
        axis.line.y = element_line(size = 1),
        axis.ticks.y = element_line(size = 1))
qc_plot
```


In general, the samples share very similar distributions across quality 
control metric - there are a few outlier cells (see nCount_RNA). However, we 
note that the nCount_RNA and nFeature_RNA appear to have bimodal
distributions. There is also a long tail of cells with unusually high 
percent.mt values (mean of `r round(mean(svz$percent.mt), 1)`%). These cells
collectively should be considered suspect and removed from downstream.  

```{r set_qc_filter}
umi_lo_threshold <- 1000
feat_hi_threshold <- 6000 # removes 2 cells
percent_mt_threshold <- 25

bad_cells <- svz$nCount_RNA < umi_lo_threshold | 
  svz$nFeature_RNA > feat_hi_threshold |
  svz$percent.mt > percent_mt_threshold
table(bad_cells)
svz$outlier <- bad_cells
```


Re-examine the disitrbution of cells colored by whether considered outlier:

```{r plot_qc_filter}
meta_vars <- c('orig.ident','outlier','nFeature_RNA','nCount_RNA', 'percent.mt')
qc_plot <- svz@meta.data[meta_vars] %>%
  reshape2::melt(id.vars = c('orig.ident','outlier')) %>%
  ggplot(mapping = aes(x = orig.ident, y = value)) +
  geom_violin(scale = 'width') +
  ggbeeswarm::geom_quasirandom(mapping = aes(color = ifelse(outlier, 'True','False')), size = 0.35) +
  scale_color_manual(values = c('True' = 'red', 'False' = 'black')) +
  facet_wrap(. ~ variable, scales = 'free_y') +
  labs(title = 'Quality control metrics') +
  theme(axis.title = element_blank(),
        plot.title = element_text(size = 18, face = 'bold'),
        axis.line.y = element_line(size = 1),
        axis.ticks.y = element_line(size = 1),
        legend.key = element_rect(fill = NA),
        legend.position = 'bottom') +
  guides(color = guide_legend(title = 'Is outlier?',
                              override.aes = list(size = 3)))
qc_plot
```

Filter the cells:  

```{r count_qc_filter}
table('High quality cells:' = !svz$outlier,
      'Injury time-point:' = svz$orig.ident)
```

Number of high-quality cells: `r sum(!svz$outlier)`

```{r apply_qc_filter}
svz <- svz[,!svz$outlier]
svz@meta.data %>%
  ggplot(mapping = aes(x = orig.ident, y = ))
saveRDS(object = svz, file = '../data/SVZ_qcfiltered.rds')
```

```{r}
svz <- readRDS(file = '../data/SVZ_qcfiltered.rds')
time_cols <- c('SCtrl' = 'grey50',
               'S1d' = 'indianred',
               'S7d' = 'dodgerblue')

feats <- c('nCount_RNA','nFeature_RNA','percent.mt')
vln <- vector(mode = 'list', length = length(feats))
names(vln) <- feats
for (f in feats) {
  # fmin <- min(svz@meta.data[[f]])
  fmax <- max(svz@meta.data[[f]])
  ndigits <- floor(log10(fmax)) + 1
  limit_range <- seq(1:9)*10^(ndigits-1)
  limit_max <- limit_range[which(limit_range > fmax)[1]]
  if ((limit_max / 10^ndigits) <= 0.5) {
    flimits <- seq(0, limit_max, by = 5*10^(ndigits-2))
  } else {
    flimits <- seq(0, limit_max, by = 10^(ndigits - 1))
  }

  vln[[f]] <- svz@meta.data[c('orig.ident', f)] |>
    mutate(orig.ident = factor(x = orig.ident, 
                               levels = c('SCtrl','S1d','S7d'))) |>
    ggplot() + 
    geom_violin(mapping = aes_string(x = 'orig.ident',
                                     y = f,
                                     fill = 'orig.ident')) +
    scale_fill_manual(values = time_cols) +
    scale_y_continuous(limits = c(0, NA),
                       breaks = flimits) +
    labs(title = f) +
    theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
          axis.text.y = element_text(size = 12),
          axis.title = element_blank(),
          strip.text = element_text(size = 12),
          panel.background = element_rect(color = 'black', fill = NA),
          panel.border = element_rect(color = 'black', fill = NA),
          axis.ticks.x = element_blank(),
          strip.background = element_blank(),
          legend.position = 'none')
}
vln_plot <- Reduce(f = `/`, x = vln)

# Labeled
svg(filename = paste0(results_out, 'QC_vln.svg'), height = 6, width = 3)
vln_plot
dev.off()

# Without labels
vln_plotX <- vln_plot & theme(title = element_blank(),
                              axis.text.x = element_blank())
svg(filename = paste0(results_out, 'QC_vln_unlabeled.svg'), 
    height = 4.5, width = 3)
vln_plotX
dev.off()
ggsave(filename = paste0(results_out, 'QC_vln_unlabeled.tiff'),
       plot = vln_plotX, device = 'tiff', height = 4.5, width = 3)
```


```{r sessionInfo}
sessionInfo()
```


