---
title: "2_SVZ_Clustering"
author: "James Choi"
date: "Last compiled on: `r Sys.Date()`"
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.align='center', 
                      tidy.opts=list(width.cutoff=60), tidy=TRUE,
                      results='hold')
```

```{r directory_setup}
if(!grepl('scripts', getwd())) {setwd('scripts')}
results_out <- '../results/2_SVZ_Clustering/'
dir.create(path = results_out)
```

```{r libraries}
require('Seurat')
require('ggplot2')
require('dplyr')
set.seed(123)
```

```{r params_utils}
shuffle_rows <- function(x) return(x[sample(1:nrow(x), size = nrow(x)),])
firstup <- function(x) {
  x <- tolower(x)
  substr(x, 1, 1) <- toupper(substr(x, 1, 1))
  return(x)
}
remove_anno <- theme(panel.background = element_rect(color = NA, fill = NA),
                     panel.border = element_rect(color = NA, fill = NA),
                     panel.grid = element_blank(),
                     axis.title = element_blank(),
                     axis.text = element_blank(),
                     axis.ticks = element_blank(),
                     legend.title = element_blank(),
                     legend.key = element_blank(),
                     legend.background = element_blank())
svz_cols <-  c('NSC' = '#f58231',
               'Microglia' = '#4363d8',
               'Unknown/Ependymal' = '#ffe119')
```

```{r load_data}
# This Seurat object contains all cells passed quality control.  
svz <- readRDS(file = '../data/SVZ_qcfiltered.rds')
```

```{r variable_features, fig.height=2.75, fig.width=5}
DefaultAssay(svz) <- 'RNA'
svz <- NormalizeData(svz, verbose = FALSE)
svz <- CellCycleScoring(
  object = svz,
  s.features = intersect(firstup(cc.genes$s.genes), rownames(svz)),
  g2m.features = intersect(firstup(cc.genes$g2m.genes), rownames(svz))
)
svz$CC.difference <- svz$S.Score - svz$G2M.Score
svz <- SCTransform(svz, 
                   assay = 'RNA', 
                   variable.features.n = 2000, 
                   vars.to.regress = 'CC.difference', 
                   verbose = FALSE)
VariableFeaturePlot(svz) + theme_bw()
```

```{r pca_elbow, fig.height=2.5, fig.width=4}
svz <- RunPCA(svz, npcs = 40, verbose = FALSE)
ElbowPlot(svz, ndims = 40) + theme_bw()
```


```{r pca_iteration, fig.height=6, fig.width=17, fig.cap="UMAPs using varying number of PCs"}
test_pcs <- 10:19
umaps <- vector(mode = "list", length = length(test_pcs))
for (i in 1:length(test_pcs)) {
  svz <- FindNeighbors(svz, dims = 1:test_pcs[i], verbose = FALSE)
  svz <- FindClusters(svz, resolution = 0.8, verbose = FALSE)
  svz <- RunUMAP(svz, dims = 1:test_pcs[i], verbose = FALSE)
  umaps[[i]] <- DimPlot(svz, pt.size = 1, label = TRUE, label.size = 6) +
    theme_bw() + labs(title = paste("# PCs:", test_pcs[i])) +
    theme(legend.position = 'none',
          axis.text = element_blank(),
          axis.ticks = element_blank())
}
umaps <- cowplot::plot_grid(plotlist = umaps, ncol = length(test_pcs)/2)
umaps
```

UMAPs with varying number of PCs and high resolution shows that the small
group of cells successfully clusters out using lower number of PCs (10-12). 
To better choose PCs, we look at gene loadings for each component:  


```{r dimloadings, fig.height=4.5, fig.width=8}
VizDimLoadings(svz, dims = 9:12, nfeatures = 30, ncol = 4)
```

We also inspect the distrbution of cells along thesse components:  

```{r pca_plots, fig.height=3.5, fig.width=12}
p1 <- DimPlot(svz, pt.size = 2, reduction = 'pca', dims = c(10,11))
p2 <- DimPlot(svz, pt.size = 2, reduction = 'pca', dims = c(10,12))
p3 <- DimPlot(svz, pt.size = 2, reduction = 'pca', dims = c(11,12))
p1 | p2 | p3
```


```{r svz_umap, fig.height=3.5, fig.width=4.5}
svz <- FindNeighbors(svz, dims = 1:10, verbose = FALSE)
svz <- RunUMAP(svz, dims = 1:10, verbose = FALSE)
svz <- FindClusters(svz, verbose = FALSE)
svz$default_cluster <- svz$seurat_clusters
Idents(svz) <- 'default_cluster'
svz_default_umap <- FetchData(svz, vars = c('default_cluster', 'UMAP_1', 'UMAP_2')) %>%
  shuffle_rows() %>%
  ggplot(mapping = aes(x = UMAP_1, y = UMAP_2)) + 
  geom_point(mapping = aes_string(fill = 'default_cluster'),
             color = 'black', pch = 21, size = 3) +
  theme_bw() +
  theme(legend.title = element_blank())
svz_default_umapX <- svz_default_umap +
  remove_anno +
  theme(legend.position = 'none')
ggsave(filename = paste0(results_out, 'SVZ_defaultClusters_umap.svg'),
       plot = svz_default_umap, device = 'svg', height = 3.5, width = 4.5)
ggsave(filename = paste0(results_out, 'SVZ_defaultClusters_umap_unlabeled.svg'),
       plot = svz_default_umapX, device = 'svg', height = 3.5, width = 3.5)
svz_default_umapX
```

Identify marker genes for each cluster to help classify as MG or NSC.  

```{r marker_table}
Idents(svz) <- 'default_cluster'
deg <- FindAllMarkers(svz, assay = 'RNA', only.pos = TRUE)
top_genes <- deg %>%
  group_by(cluster) %>%
  top_n(n = 3, wt = -p_val_adj) %>%
  ungroup()
knitr::kable(x = top_genes, caption = 'Top 3 DE genes per cluster.')
write.csv(deg, file = paste0(results_out, 'SVZ_defaultCluster_DEG.csv'))
```


```{r deg_heatmap, fig.height=7.75, fig.width=5.25, fig.cap='Heatmap of top 5 markers per cluster'}
top_genes <- deg %>%
  group_by(cluster) %>%
  top_n(n = 5, wt = -p_val_adj) %>% 
  .[['gene']] %>% 
  unique()
sig_genes <- deg %>%
  filter(p_val_adj < 0.001) %>% 
  .[['gene']] %>%
  unique()

Idents(svz) <- 'default_cluster'
DefaultAssay(svz) <- 'SCT'
avg_exp <- ScaleData(object = svz[['SCT']]@data, features = top_genes)
deg_heatmap <- cbind(data.frame(t(avg_exp)), 'ident' = svz@active.ident) %>%
  reshape2::melt(id.vars = 'ident') %>%
  group_by(ident, variable) %>%
  summarise(avg_exp = mean(value)) %>%
  mutate(variable = factor(variable, levels = rev(levels(variable)))) %>%
  ggplot(mapping = aes(x = ident, y = variable, fill = avg_exp)) +
  geom_tile(color = 'black') +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradient2(low = "#053061",
                       high = "#B2182B",
                       limits = c(NA, 3),
                       na.value ="#B2182B") +
  theme(axis.text.x = element_text(angle = 0, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12)) +
  guides(fill = guide_colorbar(
    title = 'Expression\nz-score',
    barwidth = 1.25, 
    frame.colour = 'black', 
    frame.linewidth = 0.5, 
    ticks.colour = 'black',
    ticks.linewidth = 0.5))
deg_heatmapX <- deg_heatmap + 
  remove_anno +
  theme(legend.position = 'none')
ggsave(filename = paste0(results_out, 'SVZ_defaultCluster-deg_heatmap.svg'),
       plot = deg_heatmap, device = 'svg', height = 7.75, width = 5.25)
ggsave(filename = paste0(results_out, 'SVZ_defaultCluster-deg_heatmap_unlabeled.svg'),
       plot = deg_heatmapX, device = 'svg', height = 7.75, width = 4.75)
deg_heatmap
```

Generate violin plots of common microglia and myeloid-associated genes. Gene
products of some of these genes are commonly targeted epitopes for antibodies
(e.g. Adgre == F4/80, Ptprc = CD11b, Mrc1 = CD206).  

```{r myeloid_marker_umap, fig.height=8, fig.width=13.5, fig.cap='Myeloid gene expression overlaid on SVZ UMAPs.'}
myeloid_genes <- unique(c('P2ry12','Tmem119','Cx3cr1','Csf1r','Adgre1','Ptprc','Ly6c2','Ccr2','Mrc1','Lyve1','Mki67','Top2a', 'Fcrls', 'Siglech', 'Olfml3', 'Gpr34', 'Trem2', 'Hexb'))
Idents(svz) <- 'default_cluster'
DefaultAssay(svz) <- 'RNA'
myeloid_umap <- FeaturePlot(
  object = svz,
  features = myeloid_genes,
  cols = c('grey85','blue3'),
  slot = 'data',
  combine = FALSE,
  order = TRUE
)
myeloid_umap <-  lapply(
  X = myeloid_umap,
  FUN = function(x) {
    x +
      theme_bw() +
      guides(color = guide_colorbar(frame.colour = 'black',
                                    ticks.colour = 'black',
                                    barheight = unit(2, 'cm'),
                                    barwidth = unit(0.5, 'cm')))
  }
)
myeloid_umapX <- lapply(
  X = myeloid_umap,
  FUN = function(x) {
    x <- x +
      remove_anno +
      theme(axis.line = element_blank(),
            legend.position = 'none')
    return(x)
  }
)
myeloid_umap <- cowplot::plot_grid(plotlist = myeloid_umap)
myeloid_umapX <- cowplot::plot_grid(plotlist = myeloid_umapX)
ggsave(filename = paste0(results_out, 'SVZ_myeloid-genes_umap.svg'),
       plot = myeloid_umap, device = 'svg', height = 8, width = 13)
ggsave(filename = paste0(results_out, 'SVZ_myeloid-genes_umap_unlabeled.svg'),
       plot = myeloid_umapX, device = 'svg', height = 8, width = 10)

myeloid_umap
```

```{r}
myeloid_genes <- unique(c('P2ry12','Tmem119','Cx3cr1','Csf1r','Adgre1','Ptprc','Ly6c2','Ccr2','Mrc1','Lyve1','Mki67','Top2a', 'Fcrls', 'Siglech', 'Olfml3', 'Gpr34', 'Trem2', 'Hexb'))
Idents(svz) <- 'default_cluster'
DefaultAssay(svz) <- 'RNA'
myeloid_umap <- lapply(
  X = myeloid_genes,
  FUN = function(x) {
    tmp <- FetchData(svz, c(x, 'UMAP_1', 'UMAP_2'))
    tmp[order(tmp[[x]]),] %>% 
      ggplot(mapping = aes(x = UMAP_1, y = UMAP_2)) +
      geom_point(mapping = aes_string(fill = x),
                 pch = 21, size = 3) +
      theme_bw() +
      scale_fill_gradient(low = 'grey90', high = 'blue2') +
      guides(fill = guide_colorbar(frame.colour = 'black',
                                    ticks.colour = 'black',
                                    barheight = unit(2, 'cm'),
                                    barwidth = unit(0.5, 'cm'))) +
      remove_anno +
      theme(axis.line = element_blank(),
            legend.position = 'none')
  }
)
myeloid_umap[[1]]
for (i in 1:length(myeloid_umapX)) {
  ggsave(
    filename = paste0(results_out, 'SVZ_myeloid-', 
                      myeloid_genes[i], '_umap_unlabeled.tiff'),
    plot = myeloid_umap[[i]], device = 'tiff', height = 4, width = 4)
}
```

Plot UMAPs of NSC gene expression to better visualize spread of detection
over all cells.  

```{r nsc_marker_umap, fig.height=11, fig.width=13.5, fig.cap='NSC gene expression overlaid on SVZ UMAPs.'}
nsc_genes <- unique(c('Gfap','Sox2','Dcx','Ascl1','Sox9','Id3','Id4','Foxm1','Egfr', 'Mki67','Top2a','Dlx1','Dlx2','Fos','Dcx', 'Nfib', 'Stmn2', 'Sox11', 'Gad1','Slc1a3', 'Fgfr3', 'Bcan', 'Clu', 'Gja1','Slc1a2','Mki67', 'Pcna', 'Dlx2', 'Hnrnpab','Hmgn2','Ascl1', 'Egfr', 'Lima1', 'Rpl12', 'Mfge8','Pcna','Rpl41'))

Idents(svz) <- 'default_cluster'
DefaultAssay(svz) <- 'RNA'
nsc_umap <- FeaturePlot(
  object = svz,
  features = nsc_genes,
  cols = c('grey85','blue3'),
  slot = 'data',
  combine = FALSE,
  order = TRUE
)
nsc_umap <-  lapply(
  X = nsc_umap,
  FUN = function(x) {
    x +
      theme_bw() +
      guides(color = guide_colorbar(frame.colour = 'black',
                                    ticks.colour = 'black',
                                    barheight = unit(2, 'cm'),
                                    barwidth = unit(0.5, 'cm')))
  }
)
nsc_umapX <- lapply(
  X = nsc_umap,
  FUN = function(x) {
    x <- x +
      remove_anno +
      theme(axis.line = element_blank(),
            legend.position = 'none')
    return(x)
  }
)

nsc_umap1 <- cowplot::plot_grid(plotlist = nsc_umap[1:16])
nsc_umap1X <- cowplot::plot_grid(plotlist = nsc_umapX[1:16])
nsc_umap2 <- cowplot::plot_grid(plotlist = nsc_umap[17:length(nsc_genes)])
nsc_umap2X <- cowplot::plot_grid(plotlist = nsc_umapX[17:length(nsc_genes)])

ggsave(filename = paste0(results_out, 'SVZ_nsc-genes1_umap.svg'),
       plot = nsc_umap1, device = 'svg', height = 9.5, width = 11.5)
ggsave(filename = paste0(results_out, 'SVZ_nsc-genes1_umap_unlabeled.svg'),
       plot = nsc_umap1X, device = 'svg', height = 9.5, width = 9)
ggsave(filename = paste0(results_out, 'SVZ_nsc-genes2_umap.svg'),
       plot = nsc_umap2, device = 'svg', height = 9.5, width = 11.5)
ggsave(filename = paste0(results_out, 'SVZ_nsc-genes2_umap_unlabeled.svg'),
       plot = nsc_umap2X, device = 'svg', height = 9.5, width = 9)

nsc_umap1 | nsc_umap2
```

We can also compare the data with a previously published study on NSCs vs 
ependymal cells in V-SVZ by [Shah, Stratton, et al.](https://doi.org/10.1016/j.cell.2018.03.063).
Specifically, we refer to Figure 3. Acta2 and Foxj1 are unique to ependymal
cells. Interestingly, they also note a small Flt1(+) population that we also
identified in the initial clustering of all SVZ cells.

```{r misc_genes, fig.height=5, fig.width=9}
ependymal_genes <- c('Foxj1','Cfap126','Ccdc153', 'Tmem212', 'Flt1', 'Pecam1')
Idents(svz) <- 'default_cluster'
DefaultAssay(svz) <- 'RNA'
ependymal_umap <- FeaturePlot(
  object = svz,
  features = ependymal_genes,
  cols = c('grey85','blue3'),
  slot = 'data',
  combine = FALSE,
  order = TRUE
)
ependymal_umap <-  lapply(
  X = ependymal_umap,
  FUN = function(x) {
    x +
      theme_bw() +
      guides(color = guide_colorbar(frame.colour = 'black',
                                    ticks.colour = 'black',
                                    barheight = unit(2, 'cm'),
                                    barwidth = unit(0.5, 'cm')))
  }
)
ependymal_umapX <- lapply(
  X = ependymal_umap,
  FUN = function(x) {
    x <- x +
      remove_anno +
      theme(axis.line = element_blank(),
            legend.position = 'none')
    return(x)
  }
)
ependymal_umap <- cowplot::plot_grid(plotlist = ependymal_umap)
ependymal_umapX <- cowplot::plot_grid(plotlist = ependymal_umapX)
ggsave(filename = paste0(results_out, 'SVZ_ependymal-genes_umap.svg'),
       plot = ependymal_umap, device = 'svg', height = 5, width = 9)
ggsave(filename = paste0(results_out, 'SVZ_ependymal-genes_umap_unlabeled.svg'),
       plot = ependymal_umapX, device = 'svg', height = 5, width = 7)
ependymal_umap
```


```{r assign_celltype}
svz$celltype <- plyr::mapvalues(
  x = svz$default_cluster,
  from = 0:8,
  to = c('Microglia', 'Microglia', 'NSC', 'NSC', 'Microglia', 'Microglia', 
         'NSC', 'Microglia', 'Unknown/Ependymal')
)
```

```{r celltype_umap}
feat <- 'celltype'
svz_umap <- FetchData(svz, vars = c(feat, 'UMAP_1', 'UMAP_2')) %>%
  shuffle_rows() %>% 
  ggplot(mapping = aes(x = UMAP_1, y = UMAP_2)) + 
  geom_point(mapping = aes_string(fill = feat),
             color = 'black', pch = 21, size = 4) +
  scale_fill_manual(values = c('NSC' = '#f58231',
                               'Microglia' = '#4363d8',
                               'Unknown/Ependymal' = '#ffe119'))

# With legend
svz_umap
svg(filename = paste0(results_out, 'SVZ_celltype_umap.svg'),
    height = 3.5, width = 5.5)
svz_umap
dev.off() 

# Without legend
svz_umapX <- svz_umap +
  remove_anno + 
  theme(legend.position = 'none')
svg(filename = paste0(results_out, 'SVZ_celltype_umap_unlabeled.svg'),
    height = 3.5, width = 4)
svz_umapX
dev.off()
# TIFF
tiff(filename = paste0(results_out, 'SVZ_celltype_umap_unlabeled.tiff'),
     height = 3.5, width = 4, units = 'in', res = 440)
svz_umapX
dev.off() 
```

```{r celltype_heatmap, fig.height=8, fig.width=4}
all_genes <- c(myeloid_genes, nsc_genes, ependymal_genes)
expr_data <- ScaleData(svz[['RNA']]@data[all_genes,], scale.max = 4) %>% 
  Matrix::t() %>% 
  cbind(svz@meta.data['celltype']) %>%
  group_by(celltype) %>%
  summarise(across(.cols = all_of(all_genes), .fns = mean)) %>%
  tibble::column_to_rownames(var = 'celltype') %>%
  t()
svz_heatmap <- ComplexHeatmap::Heatmap(
  matrix = expr_data,
  clustering_method_columns = 'ward.D2',
  clustering_method_rows = 'ward.D2',
  # col = heatmap_cols,
  border = TRUE,
  clustering_distance_rows = 'pearson',
  clustering_distance_columns = 'pearson',
  column_dend_height = unit(x = 15, units = 'mm'),
  row_dend_width = unit(x = 15, units = 'mm'),
  column_names_rot = 45,
  rect_gp = grid::gpar(col = 'black'),
  row_names_gp = grid::gpar(fontface = 'italic'),
  heatmap_legend_param = list(
    title = 'z-score',
    title_gp = grid::gpar(fontsize = 12),
    title_position = 'leftcenter-rot',
    labels = c('Low', 0, 'High'),
    at = c(min(expr_data), 0, max(expr_data)),
    labels_gp = grid::gpar(fontsize = 10),
    legend_height = unit(2.5, units = 'cm'),
    grid_width = unit(0.5, units = 'cm'),
    border = 'black',
    title_gap = unit(1, units = 'cm'),
    direction = 'vertical')
)
svg(filename = paste0(results_out, 'SVZ_celltype-marker_heatmap.svg'),
    height = 8, width = 4)
svz_heatmap_print <- cowplot::as_gtable(
  plot = grid::grid.grabExpr(
    expr = ComplexHeatmap::draw(
      object = svz_heatmap,
      padding = grid::unit(c(2,8,2,2),units = 'mm')
    )
  )
)
plot(svz_heatmap_print)
dev.off()
```

```{r save_data}
saveRDS(svz, file = '../data/svz.rds')
```

```{r session_info}
sessionInfo()
```

